Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-03 pecomplexity.tex
update results section

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:cache}, and chaos places fundamental limits on
predictability.
%
\begin{figure}[htbp]
   \centering
   \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
   % where an .eps filename suffix will be assumed under latex,
   % and a .pdf suffix will be assumed for pdflatex
   \caption{A small snippet of the L2 cache miss rate of {\tt
       col\_major}, a three-line C program that repeatedly initializes
     a matrix in column-major order, running on an Intel Core
     Duo\textsuperscript{\textregistered}-based machine.  Even this
     simple program exhibits chaotic performance dynamics.}
   \label{fig:cache}
 \end{figure}

@@

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:cache}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.
\begin{figure}
  \centering
    \includegraphics[width=0.5\textwidth]{figs/3dcolcache}
    \caption{A 3D projection of a delay-coordinate embedding of the trace
from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
}
\label{fig:embedding}
\end{figure}
%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \verb|col_major| L2 cache miss rate time series
$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
$y(t+2\tau)$ on the third, and so on.
%% ...down to here.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism\footnote{A deeper analysis of
  Figure~\ref{fig:embedding}---as alluded to on the previous
  page---supports that diagnosis, confirming the presence of a chaotic
  attractor in these cache-miss dynamics, with largest Lyapunov
  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
structure can also be used to build a forecast model.

@@

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\cite{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes.

@@

The relationship between prediction accuracy and the weighted permutation
entropy (WPE) is much as we conjectured: performance traces with high WPE ---
those whose temporal complexity is high, in the sense that the entropy
generation per value is approximately equal to the total entropy per value thus
little information is being propagated forward in time --- are indeed harder to
predict using the simple deterministic forecast model described in the previous
section. See Fig.~\cite{fig:pred_vs_ent} for the distribution. The effects of
changing the word length $n$ are also interesting: using a longer word length
generally lowers the WPE --- a natural consequence of finite-length data --- but
the falloff is less rapid in some traces than in others, suggesting that those
values are closer to the theoretical asymptote that exists for perfect data. The
persistent WPE values of 0.5--0.6 for the {\tt col\_major} trace are consistent
with dynamical chaos, further corroborating the results of~\cite{mytkowicz09}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
  \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
  the word length used is $5$.}
  \label{fig:lma_pred_vs_ent}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
  \caption{The MASE of ARIMA vs weighted permutation entropy. For each of these,
  the word length used is $5$.}
  \label{fig:arima_pred_vs_ent}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $5$.}
  \label{fig:pred_vs_ent}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
  \caption{The MASE values for LMA against ARIMA. The dashed line is the
  identity.}
  \label{fig:lma_vs_arima}
\end{figure}

@@

The results presented here suggest that permutation entropy---a
ordinal calculation of forward information transfer in a time
series---is an effective metric for predictability of computer
performance traces. Experimentally, traces with a persistent PE
$\gtrapprox 0.97$ have a natural level of complexity that may
overshadow the inherent determinism in the system dynamics, whereas
traces with PE $\lessapprox 0.7$ seem to be highly predictable (viz.,
at least an order of magnitude improvement in nRMSPE).

>>>>

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:cache}, and chaos places fundamental limits on
predictability.
%
% \begin{figure}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
%    % where an .eps filename suffix will be assumed under latex,
%    % and a .pdf suffix will be assumed for pdflatex
%    \caption{A small snippet of the L2 cache miss rate of {\tt
%        col\_major}, a three-line C program that repeatedly initializes
%      a matrix in column-major order, running on an Intel Core
%      Duo\textsuperscript{\textregistered}-based machine.  Even this
%      simple program exhibits chaotic performance dynamics.}
%    \label{fig:cache}
%  \end{figure}

@@

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:cache}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.
% \begin{figure}
%   \centering
%     \includegraphics[width=0.5\textwidth]{figs/3dcolcache}
%     \caption{A 3D projection of a delay-coordinate embedding of the trace
% from Figure~\ref{fig:cache} with a delay ($\tau$) of 100,000 instructions.
% }
% \label{fig:embedding}
% \end{figure}
%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \verb|col_major| L2 cache miss rate time series
$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
$y(t+2\tau)$ on the third, and so on.
%% ...down to here.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism\footnote{A deeper analysis of
  Figure~\ref{fig:embedding}---as alluded to on the previous
  page---supports that diagnosis, confirming the presence of a chaotic
  attractor in these cache-miss dynamics, with largest Lyapunov
  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
structure can also be used to build a forecast model.

@@

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes.

@@

 \section{Results} % (fold)
 \label{sec:results}

The relationship between prediction accuracy and the weighted permutation
entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in the prior
section. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we can explain. Furthermore, we find
that the relationship between the two is roughly linear. The single outlier,
{\tt SVD} regime 1, does not fit the trend due to a weakness in the WPE: namely
that in the absence of any large features, as all the other regimes have, the
WPE effectively falls back to the standard PE for which the noisy behavior
drives toward 1.0.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$.}
  \label{fig:pred_vs_ent}
\end{figure}

In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA and
the naive method. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. This however does not contradict our hypothesis,
it just means that ARIMA is ill-suited to the time series under investigation.
To see this more clearly, consider Figure~\ref{fig:lma_vs_arima}. Since larger
MASE values constitute worse predictions, ARIMA performed worse than LMA in all
traces except {\tt SVD$_1$}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
    the word length used is $6$.}
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
    these, the word length used is $6$.}
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
    \caption{The MASE values for LMA against ARIMA. The dashed line is the
    identity, delineating the traces for which either LMA or ARIMA performed
    better. All traces except those from {\tt SVD$_1$} lie above the line,
    indicating that LMA is better suited prediction method for the computer
    traces considered.}
    \label{fig:lma_vs_arima}
  \end{subfigure}
\end{figure}

 % section results (end)

@@

The results presented here suggest that permutation entropy---a ordinal
calculation of forward information transfer in a time series---is an effective
metric for predictability of computer performance traces. Experimentally, traces
with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
may overshadow the inherent determinism in the system dynamics, whereas traces
with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
corroborating the results of~\cite{mytkowicz09}.
